---
title: "How much of a shared subspace is carried by group structure?"
subtitle: "Dolors Pelegrí-Sisó and Juan R. Gonzalez"
author: |
   Institute for Global Health (ISGlobal), Barcelona, Spain
   Bioinformatics Research Group in Epidemiolgy (BRGE)
   https://brge.isglobal.org
date: "`r Sys.Date()`"
package: "`r pkg_ver('mgcca')`"
output:
  BiocStyle::html_document:
    number_sections: true
    toc: yes
    fig_caption: yes
  BiocStyle::pdf_document:
    number_sections: true
    toc: yes
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Reliability of the shared subspace}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 3.6)
```

# Introduction

Having estimated a shared subspace with `mgcca()`, two questions arise before any
of it is interpreted.

The first is whether a quantity you care about -- a phenotype, an exposure, a
score -- actually lies in that subspace, and if it does, how much of that
alignment leans on a **grouping** in your design: batch, plate, centre, cohort,
run. An association that disappears once the grouping is accounted for is a
different scientific object from one that survives it, and nothing in the fit
itself distinguishes them.

The second is whether the subspace is **reproducible**: would you have estimated
the same directions from a different sample, or from data without one particular
group?

This vignette covers the three functions that answer them:

| function | question |
|---|---|
| `mgcca_sensitivity()` | does the query align with the subspace, and how much of that is carried by the grouping? |
| `mgcca_block_profile()` | which blocks carry the sensitivity? |
| `mgcca_stability()` | is the subspace reproducible under resampling? |

## Reading the numbers without the algebra {#reading}

### Two words used throughout

**The shared subspace** is just the handful of axes the tables agree on -- the
common signal running through your proteins, transcripts and metabolites at once.
Everything below is about how a variable of yours relates to *those axes*.

**A grouping** is any variable that puts each individual into one of a few groups.
That is all it means. It might be:

* **technical** -- the plate a sample was run on, the sequencing batch, the
  recruiting centre, the week of processing;
* **biological or clinical** -- disease stage, tissue, treatment arm, cohort.

⚠️ **The functions treat both the same way; you must not.** The arithmetic is
identical, but a result that leans on *plate* and a result that leans on *disease
stage* mean opposite things. Which one you have is something only you know, and it
is the single judgement this vignette keeps asking you to make. Where the text
says "grouping" without qualifying it, it means whichever one you supplied.

### What "alignment" actually means

`T`, the alignment, is best read as **the fraction of your variable that the
shared omic axes can reproduce**. It behaves like a proportion of variance
explained: 0 means the integrated signal cannot account for your variable at all,
1 would mean it accounts for it entirely. (Strictly it is that proportion measured
in the weighting the estimator uses for individuals missing from some tables, but
nothing below depends on the distinction.)

So a statement like *"BMI has an alignment of 0.30"* reads as: **about a third of
the variation in BMI is echoed in the axes the omic tables agree on.**

⚠️ **It has no conventional threshold.** Unlike a correlation, you cannot say
"above 0.3 is meaningful". Its floor depends on how many axes you kept and how
many individuals you have, which is why every example here includes a **noise
control**: a variable made of pure random numbers, run through the same machinery.
That row tells you what "nothing at all" looks like *in your own analysis*, and
your real variable is read against it.

### The four numbers

| number | what it tells you | a concrete example |
|---|---|---|
| `T` | how much of your variable the integrated omic signal accounts for | BMI at `T = 0.30` and smoking at `T = 0.02`: the omics have a lot to say about BMI and essentially nothing about smoking |
| `S_total` | how **fragile** that reading is -- how much it would shift if the measurements were slightly different (a re-run, a drifting platform, a different normalisation) | two proteins both at `T = 0.20`, one with a small `S_total` and one with a large one: the first would read the same tomorrow, the second might not |
| `S_between` | the part of that fragility that follows your **grouping** -- batch, plate, centre, cohort, tumour stage | of everything that could shift the BMI reading, how much of it is "which plate the sample was run on" |
| `R_between` | the share: `S_between / S_total` | `R_between = 0.05`: the reading barely depends on the grouping. `R_between = 0.60`: most of what holds it up is the group structure -- if your groups were plates rather than disease stages, that is a warning |

And the practical point, which is the whole reason the layer exists:

> **Two variables with the same `T` are not the same finding** if one has a high
> `R_between` and the other does not. The fit alone gives you no way to tell them
> apart; this is what does.

### In what order to look at things

There is a sequence, and skipping to the middle of it is how results get
overinterpreted.

**Step 1 -- is the subspace itself trustworthy?** Run `mgcca_stability()` first. If
the axes move substantially when you resample your own participants, nothing that
follows should be interpreted: you would be reading a structure that a slightly
different sample would not have produced. Section \@ref(is-the-subspace-reproducible)
shows one trustworthy case and one that is not.

**Step 2 -- does the omic signal know about your variable at all?** Read `T`
against the noise control. If it is not clearly above it, stop: there is nothing
to explain.

**Step 3 -- what is holding it up?** Now `R_between` and the per-block profile.

```{r flow, echo=FALSE, fig.width=7.6, fig.height=4.8, fig.cap="The order matters: each step only makes sense if the one above it passed."}
# Drawn with ggplot2 rather than a diagram package: ggplot2 is already a hard
# dependency, so the figure cannot quietly vanish on a machine where an optional
# one is missing.
pal <- c("#0072B2", "#D55E00", "#009E73")
BOX <- function(x, y, w, label, fill, size = 3.0, h = 1.25)
  list(ggplot2::annotate("tile", x = x, y = y, width = w, height = h, fill = fill),
       ggplot2::annotate("text", x = x, y = y, label = label, colour = "white",
                         size = size, lineheight = 1.0))
AR <- function(x0, y0, x1, y1, lab = NULL)
  c(list(ggplot2::annotate("segment", x = x0, y = y0, xend = x1, yend = y1,
           arrow = grid::arrow(length = grid::unit(0.15, "cm"), type = "closed"),
           colour = "grey45", linewidth = 0.4)),
    if (!is.null(lab)) list(ggplot2::annotate("text", x = (x0 + x1)/2 - 0.30,
           y = (y0 + y1)/2, label = lab, size = 2.7, colour = "grey30", hjust = 1)))
sp <- 2.0
ggplot2::ggplot() +
  BOX(sp, 9.0, 5.4, "1.  Is the subspace stable?\nmgcca_stability()", pal[1]) +
  AR(sp - 1.4, 8.35, -4.0, 7.75, "no") + AR(sp, 8.35, sp, 7.75, "yes") +
  BOX(-4.0, 7.1, 4.0, "STOP\nnothing below\nis interpretable", "grey55", 2.8) +
  BOX(sp, 7.1, 5.4, "2.  Is T above the\nnoise floor?", pal[1]) +
  AR(sp - 1.4, 6.45, -4.0, 5.85, "no") + AR(sp, 6.45, sp, 5.85, "yes") +
  BOX(-4.0, 5.2, 4.0, "STOP\nthe omics have\nnothing to say", "grey55", 2.8) +
  BOX(sp, 5.2, 5.4, "3.  Is R_between\nhigh?", pal[1]) +
  AR(sp - 1.4, 4.55, -1.0, 3.95, "no") + AR(sp + 1.4, 4.55, 6.4, 3.95, "yes") +
  BOX(-1.0, 3.3, 5.0, "STRONGEST\na real share, and it does\nnot lean on your groups", pal[3], 2.8) +
  BOX(6.4, 3.3, 5.0, "CHECK\nwhat IS your grouping?", pal[2], 2.8) +
  AR(5.2, 2.65, 4.0, 2.05) + AR(7.6, 2.65, 8.8, 2.05) +
  BOX(4.0, 1.4, 4.4, "biological\n(stage, tissue)\n= a finding", pal[3], 2.7) +
  BOX(8.8, 1.4, 4.4, "technical\n(plate, batch)\n= suspect an artefact", pal[2], 2.7) +
  ggplot2::coord_cartesian(xlim = c(-6.3, 11.3), ylim = c(0.5, 9.9)) +
  ggplot2::theme_void() +
  ggplot2::labs(title = "In what order to read the numbers",
                subtitle = "each step only makes sense if the one above it passed") +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold", size = 11),
                 plot.subtitle = ggplot2::element_text(size = 9, colour = "grey35"))
```

### What each combination means

Assuming step 1 passed, this is how the two headline numbers combine:

| | `R_between` **low** | `R_between` **high** |
|---|---|---|
| **`T` clearly above noise** | The strongest position. The omics account for a real share of your variable, and that reading does not lean on your grouping. Report it, and still not as causal. | The omics account for it, **but the signal moves with your groups**. Two very different explanations fit, and the data cannot separate them -- see below. |
| **`T` near the noise level** | The integrated signal has essentially nothing to say about this variable. Nothing to interpret, and no amount of further analysis will change that. | The little alignment there is comes mostly from group structure. The weakest cell: small and group-driven. |

**The top-right cell is where judgement is needed, and it is worth being explicit
about why.** A high `R_between` means the alignment and the grouping move
together. That is what you would expect if:

* the grouping is **biological** -- disease stage, tissue, treatment arm -- and
  your variable genuinely varies along it. This is a finding.
* the grouping is **technical** -- plate, batch, centre, run date -- in which case
  you may be looking at a processing artefact dressed as biology.

**The analysis cannot distinguish these. You can, because you know what the
grouping is.** That is the honest limit, and it is also the useful part: a number
that flags "check this" is worth more than one that quietly does not.

Like the estimator, the numerical core runs in C++ over
[`BigDataStatMeth`](https://github.com/isglobal-brge/BigDataStatMeth): the block
Grams are formed out-of-core and never load a block into memory, while the
participant-space algebra -- which is `n × n` in *individuals*, not variables --
is done in RAM, where it belongs.

**What these quantities are not.** They are first-order sensitivity coefficients
of a fitted operator. They are not causal, and `mgcca_sensitivity()` is not a
psychometric reliability coefficient. The word *reliability* names the framework;
the functions take the narrower mathematical names.

# Installation

`mgcca` is installed from Bioconductor:

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("mgcca")
```

```{r load_library}
library(mgcca)
```

# A self-contained example

The `cardiovascular` data set ships with the package: three tables that do not
share the same individuals.

```{r cardio_fit}
data(cardiovascular)
X <- list(methylation = as.matrix(X1),
          clinical    = as.matrix(X2),
          other       = as.matrix(X3))
X <- lapply(X, function(m) { storage.mode(m) <- "double"; m })

h5file <- tempfile(fileext = ".h5")
fit <- mgcca(X, filename = h5file, nfac = 2, method = "solve", scores = TRUE)
fit
```

## Does the omic signal know anything about my variable?

A query is any participant-level vector or matrix. It must carry **individual
identifiers**: participants are matched by name, never by position. That is not
fussiness. A misaligned query returns a perfectly plausible number rather than an
error, and nothing in the output would let you catch it.

We ask about two clinical variables, and add pure noise as a control -- a
reference that should show almost no alignment, so the others can be read against
something.

```{r cardio_query}
# The clinical table covers 633 of the 646 individuals in the fit. A query does
# NOT have to cover everyone: those it misses simply contribute nothing, and
# `n_used` reports how many actually did.
cids <- intersect(rownames(fit$Y), rownames(X2))

# The noise floor is ESTIMATED, not taken from one draw. A single random column is
# itself a random number: it can land high or low, and reading a real variable
# against one lucky draw is how a null result becomes a finding. Five columns give
# the floor a range instead of a point.
set.seed(2026)
noise <- replicate(5, rnorm(length(cids)))
colnames(noise) <- paste0("noise", 1:5)

Q <- cbind(LDL     = X2[cids, "LDL"],
           glucose = X2[cids, "Gluc"],
           noise)
rownames(Q) <- cids

sens <- mgcca_sensitivity(fit, Q)
sens
```

`T` is the share of the query that lies in the retained subspace, bounded in
$[0,1]$. `S_total` is its first-order sensitivity: how much that alignment would
move under a structured perturbation of the blocks.

```{r cardio_read, echo=FALSE, results='asis'}
o     <- sens$overall
tl    <- o[["T"]][o$query == "LDL"]
tg    <- o[["T"]][o$query == "glucose"]
floor <- o[["T"]][grepl("^noise", o$query)]
hi    <- max(floor)
say <- function(x, nm) {
  if (x > 3 * hi)
    sprintf(paste0("**%s clears the ceiling comfortably.** The omic signal accounts ",
                   "for %.2f%% of it, against a ceiling of %.2f%%, so what it picks ",
                   "up is not something random numbers would have produced."),
            nm, 100 * x, 100 * hi)
  else if (x > hi)
    sprintf(paste0("**%s is above the ceiling, but only just** -- %.2f%% against ",
                   "%.2f%%. A gap that narrow is worth noting and not worth ",
                   "building on: treat it as suggestive, and see whether it holds ",
                   "in a second cohort before it goes in an abstract."),
            nm, 100 * x, 100 * hi)
  else
    sprintf(paste0("**%s does not clear the ceiling at all** (%.2f%% against ",
                   "%.2f%%). The integrated signal accounts for no more of it than ",
                   "it would of random numbers, so there is nothing here to explain ",
                   "and no follow-up analysis will change that."),
            nm, 100 * x, 100 * hi)
}
cat(sprintf(paste0(
  "**Reading this table.** The shared axes account for %.2f%% of the variation in ",
  "LDL and %.2f%% of glucose.\n\n",
  "Now, is that a lot or a little? On its own, neither -- the number has no ",
  "conventional threshold. That is what the five `noise` rows are for. They are ",
  "columns of random numbers put through exactly the same machinery, and they come ",
  "back between %.2f%% and %.2f%%. **A variable with no relationship whatsoever ",
  "still scores in that band**, simply because projecting anything onto a couple ",
  "of axes catches a little by chance. So that band is the ceiling that ",
  "meaningless variables reach in this analysis, and it is the yardstick.\n\n",
  "(Five columns rather than one, because a single random column is itself a ",
  "random draw: it might land at %.2f%% or at %.2f%%. Reading a real variable ",
  "against one lucky draw is exactly how a null result turns into a finding.)\n\n",
  "%s\n\n%s\n\n",
  "⚠️ What this does **not** say: nothing about methylation values being higher or ",
  "lower, nothing about direction, and nothing about causing or being caused. It ",
  "says how much of a variable's variation the integrated signal can reproduce, ",
  "and whether that is more than chance alone would give.\n"),
  100 * tl, 100 * tg, 100 * min(floor), 100 * hi,
  100 * min(floor), 100 * hi,
  say(tl, "LDL"), say(tg, "Glucose")))
```

`S_between` and `R_between` are `NA` here, and deliberately so: **no grouping was
supplied**, so the group-dependent quantities are *absent* rather than filled
with a zero that would look like a measurement. `cardiovascular` carries no batch,
centre or run variable, which is exactly the common situation in which only the
total is available. Section \@ref(a-real-grouping) uses a data set that does have
one.

```{r cardio_floor, fig.height=3.6, fig.cap="Only a query whose point sits outside the grey band is saying anything the machinery would not have produced from random numbers."}
plot(sens, type = "overall", noise = paste0("noise", 1:5))
```

**How to read this figure.** One row per variable, and the grey band is the level
the control columns reached -- what chance alone gives in *this* analysis. A point
inside the band is indistinguishable from noise; a point clear of it is not. It is
the same comparison the table above makes, but you can take it in at a glance,
and it scales: with two hundred exposures the table becomes unreadable and this
does not.

## Which data type is the finding coming from?

```{r cardio_profile}
prof <- mgcca_block_profile(sens)
prof
```

`mgcca_block_profile()` **extracts**; it does not recompute. The block-level
quantities come back from the same pass that produced the totals, so there is one
route to them. Two independent routes would eventually disagree, and the
disagreement would be silent.

Read the shares next to the absolute values. A block carrying most of a
negligible sensitivity is still carrying most of it, and the share alone would
not tell you which situation you are in.

```{r cardio_profile_plot, fig.height=4.4}
plot(sens)
```

**How to read this figure.** One panel per variable, one bar per table. The bar
height is how much that table contributes to the variable's sensitivity, so the
tallest bar names the data type the finding actually rests on. In a paper this is
the difference between *"the integrated signal relates to LDL"* and *"the relation
is essentially methylation, with the clinical table adding little"* -- and only
the second is a statement a reviewer can act on.

```{r cardio_profile_read, echo=FALSE, results='asis'}
bb <- prof$profile[prof$profile$query == "LDL", ]
top <- bb$block[which.max(bb$share_total)]
cat(sprintf(
  "Here the largest contributor to LDL is **%s** (%.0f%% of its total sensitivity).\n",
  top, 100 * max(bb$share_total)))
```

# When your study has batches, centres or stages {#a-real-grouping}

`miniACC` (from `r Biocpkg("MultiAssayExperiment")`) is a compact public data set
that offers both of the things this layer needs: individuals **missing from some
assays**, and a genuine clinical **grouping**.

```{r acc_available, echo=FALSE}
have_acc <- requireNamespace("MultiAssayExperiment", quietly = TRUE) &&
            requireNamespace("SummarizedExperiment", quietly = TRUE)
```

```{r acc_build, eval=have_acc}
library(MultiAssayExperiment)
library(SummarizedExperiment)
data("miniACC", package = "MultiAssayExperiment")

# Assays are features x samples; mgcca wants individuals x variables with the
# patient ID as row name, so each assay is renamed through the sample map and
# transposed. A few hundred variables per block keeps the vignette quick.
sm  <- sampleMap(miniACC)
use <- c(mRNA = "RNASeq2GeneNorm", protein = "RPPAArray", miRNA = "miRNASeqGene")

tabs <- lapply(use, function(a) {
  m <- assay(experiments(miniACC)[[a]])
  k <- sm[sm$assay == a, ]
  colnames(m) <- as.character(k$primary[match(colnames(m), as.character(k$colname))])
  m <- m[, !duplicated(colnames(m)), drop = FALSE]
  Xa <- t(m); storage.mode(Xa) <- "double"
  keep <- apply(Xa, 2, function(v) all(is.finite(v)) && sd(v) > 0)
  Xa[, which(keep)[seq_len(min(sum(keep), 150))], drop = FALSE]
})

sapply(tabs, dim)
length(Reduce(union, lapply(tabs, rownames)))       # union of individuals
length(Reduce(intersect, lapply(tabs, rownames)))   # present in all three
```

Of the 80 patients, only 46 appear in all three assays. The remaining 34 are
present in some and absent from others -- the setting `mgcca` exists for, and the
reason the availability weighting matters here rather than being a formality.

```{r acc_fit, eval=have_acc}
h5acc <- tempfile(fileext = ".h5")
acc <- mgcca(tabs, filename = h5acc, nfac = 2,
             method = "penalized", lambda = rep(1, 3), scores = TRUE)

ids  <- rownames(acc$Y)
cd   <- as.data.frame(colData(miniACC))
stage <- factor(as.character(cd[ids, "pathologic_stage"]))
names(stage) <- ids
table(stage, useNA = "ifany")
```

## Is my result leaning on those groups?

Now the grouped quantities are defined. We ask about tumour stage as an ordered
score, against the same noise control -- five columns of it, for the reason
spelt out below.

```{r acc_sens, eval=have_acc}
score <- setNames(as.numeric(stage), ids)
# FIVE noise columns, not one. A single random column is a DRAW from the noise
# level, not the noise level -- and on the map below, one draw lands wherever it
# lands, sometimes squarely along the direction the groups separate in. Five of
# them show the spread, which is the honest comparison.
set.seed(2026)
Q <- cbind(stage_score = score, replicate(5, rnorm(length(ids))))
colnames(Q) <- c("stage_score", paste0("noise", 1:5))
rownames(Q) <- ids

sens_acc <- mgcca_sensitivity(acc, Q, group = stage)
sens_acc
```

```{r acc_sens_read, eval=have_acc, echo=FALSE, results='asis'}
o  <- sens_acc$overall
sg <- o[o$query == "stage_score", ]; nzr <- range(o[["T"]][grepl("^noise", o$query)])
r  <- sg$R_between
band <- if (r < 0.05) "very low" else if (r < 0.20) "low" else if (r < 0.50) "moderate" else "high"
cat(sprintf(paste0(
  "**Reading this table.** The three assays together account for %.2f%% of the ",
  "variation in tumour stage, against %.2f--%.2f%% for the five random columns. ",
  "Of everything that could shift that reading, **%.1f%% follows stage itself as ",
  "a grouping** (`R_between` = %.3f) -- a %s value.\n\n",
  "\u26a0\ufe0f Note where the alignment sits relative to the noise columns: %s. ",
  "With this many patients and two retained axes, an arbitrary vector picks up a ",
  "share of the subspace by chance, and `T` alone does not distinguish the two. ",
  "What does is the grouping decomposition below, and the direction on the map.\n"),
  100 * sg[["T"]], 100 * nzr[1], 100 * nzr[2], 100 * r, r, band,
  if (sg[["T"]] > nzr[2]) "above every one of them"
  else if (sg[["T"]] < nzr[1]) "**below every one of them**"
  else "**inside their range**"))
```

**What to do with that number, whatever it had turned out to be.** This is the
part worth having in advance, because the temptation is always to decide what a
value means after seeing it.

| if `R_between` were... | what it would be saying | what to do next |
|---|---|---|
| **below ~0.05** | the reading barely depends on the grouping at all | nothing further on this axis. Report the alignment; the grouping is not what is holding it up |
| **around 0.1--0.3** | a real but minority share follows the grouping | check *which* blocks carry the between-group part (`mgcca_block_profile()`). If it concentrates in one assay, that assay's processing is the thing to look at |
| **above ~0.5** | most of what holds the reading up is group structure | treat the alignment as a statement about the groups until shown otherwise. Re-run with the grouping residualised out via `covariates =`, and see what survives |
| **high, and the grouping is technical** | plate, batch, centre, run date | this is the artefact case. Look at whether the batches are confounded with anything biological, and consider re-normalising before re-fitting |
| **high, and the grouping is biological** | stage, tissue, treatment arm | this may be exactly the finding. But confirm the biological grouping is not itself confounded with a technical one -- samples of the same stage collected in the same period is a very common way for the two to become indistinguishable |

**The case in front of us.** A high `R_between` means the alignment and the group
structure move together. That is what you would expect from a genuine biological
gradient -- patients at the same stage really do resemble each other -- and also
exactly what a batch confounded with stage would produce. **The analysis cannot
separate those two. You can, because you know how the samples were collected.**

If you cannot rule the technical explanation out, the honest options are: report
the alignment as conditional on the grouping, or re-run with the grouping
residualised out and report what is left. Both are defensible. Reporting the raw
alignment without mentioning the grouping is not.



`R_between` is the share of the sensitivity carried by the grouping. **Read it
with its two components, never alone.** A high ratio can sit on a negligible
absolute sensitivity, and a low ratio on a large one; the ratio says how the
sensitivity is distributed, not how much of it there is. That is why `S_total` and
`S_between` are printed beside it rather than being folded away.

```{r acc_plot, eval=have_acc, fig.height=4.6}
plot(sens_acc)
```

**How to read this figure.** Each bar is one assay. The height says which assay
the finding rests on; the inner bar and its percentage say how much of that
assay's contribution follows the grouping. An assay with a tall bar and a small
percentage is carrying signal that does not depend on the groups -- the most
comfortable position to be in. A tall bar with a large percentage is the one to
look at twice.

The bars are the total sensitivity per block; the inner bar and the label are the
part carried by the grouping. They are drawn overlaid rather than side by side
for a plain reason: the between-group part is routinely two orders of magnitude
shorter, so side by side it renders as an invisible sliver and the figure fails
to show the one comparison it exists for.

## Seeing it instead of inferring it

`R_between` says *how much* of the reading follows the grouping. The map view says
*in which direction*, and that is often the more persuasive of the two.

```{r acc_map, eval=have_acc, fig.height=4.8, fig.cap="Individuals on the reference axes, group centroids as diamonds, and each query as an arrow."}
plot(sens_acc, type = "map")
```

Each point is a patient placed on the axes the assays agree on; the diamonds are
the group centroids; each arrow is the direction a query represents.

**What to look for.** The dashed line is the direction the groups separate in.
If a query's arrow runs along it, the query and the grouping are describing the
same movement -- which is exactly what a high `R_between` measures. If it points
elsewhere, they are not. Arrow *lengths* mean nothing here: they are all scaled
alike, and only the angle carries information.

```{r acc_map_angles, eval=have_acc, echo=FALSE, results='asis'}
# The angles are MEASURED, not read off the picture. Two people looking at the same
# arrows will not agree to within ten degrees, and this comparison lives or dies on
# about that much.
R2 <- sens_acc$reference
Y2 <- R2$Y[, 1:2, drop = FALSE]
gv <- R2$group[rownames(Y2)]
cen <- t(vapply(split(seq_len(nrow(Y2)), gv),
                function(i) colMeans(Y2[i, , drop = FALSE]), numeric(2)))
gdir <- cen[nrow(cen), ] - cen[1L, ]
# Angle to the LINE, not the ray: the dashed line is drawn through the origin both
# ways, so a query pointing "backwards" along it agrees just as strongly.
angl <- function(u) 180 / pi * acos(pmin(1, abs(sum(u * gdir)) /
                                    sqrt(sum(u^2) * sum(gdir^2))))
W2 <- R2$directions[1:2, , drop = FALSE]
aa <- vapply(colnames(W2), function(q) angl(W2[, q]), numeric(1))
an <- aa[grepl("^noise", names(aa))]
cat(sprintf(paste0(
  "Measured from the figure's own coordinates: **tumour stage sits %.0f\u00b0 from ",
  "the dashed line**, while the five noise columns land at %.0f\u00b0, %.0f\u00b0, ",
  "%.0f\u00b0, %.0f\u00b0 and %.0f\u00b0. %s\n"),
  aa[["stage_score"]], an[1], an[2], an[3], an[4], an[5],
  if (aa[["stage_score"]] < min(an))
    paste0("Stage is closer to the group direction than **every** noise draw -- ",
           "which is the comparison that matters, and it is the one the single ",
           "number `T` failed to make.")
  else
    paste0("\u26a0\ufe0f At least one noise draw lies as close to the group ",
           "direction as tumour stage does. That is not a defect in the figure: ",
           "with this sample size a random vector can fall anywhere, which is ",
           "precisely why one control column is never enough.")))
```

**And something the numbers do not tell you.** Look at how much the ellipses
overlap. Stage has a direction in this space, but the groups are not separated
along it -- individual patients of different stages are thoroughly mixed. A
clinician takes that in immediately, and it tempers the reading in a way no single
coefficient can: a real but weak gradient, not a clean separation.

With more than two components, call `plot(..., comps = c(1, 3))` and so on. Three
dimensions have to be rotated to be read and do not survive a printed page; beyond
three there is nothing to rotate.

## Which participants is it coming from?

The per-block profile says *which data type* carries the sensitivity. There is a
second cut of exactly the same quantity: `S_total` is a sum over **individuals**
as much as over blocks, and `by_individual` reports the per-participant terms of
that sum. Same arithmetic, different axis.

```{r acc_ind_head, eval=have_acc}
bi <- sens_acc$by_individual
head(bi[bi$query == "stage_score", ], 4)
```

One row per participant, per block, per query. `S_total` is that participant's
contribution to that block's sensitivity; `share_total` is the same thing as a
fraction of the block total, so within one query and block the shares add to one.

⚠️ **Key this table by `query`, `id` and `block`, never by row position.** Row
order is not part of the contract. It is the same discipline the query itself is
held to, and for the same reason: a positional read produces a plausible number
rather than an error.

⚠️ **`S_between` is a property of the group, not of the person.** The
between-group part of the sensitivity is built from group means, so every member
of a group carries the identical value -- here twelve distinct numbers for four
stages times three assays, repeated across the participants in each:

```{r acc_ind_between, eval=have_acc}
length(unique(bi$S_between[bi$query == "stage_score"]))   # 4 stages x 3 assays
```

So read `S_total` per participant, and `S_between` as the group's contribution
carried in each of its members' rows. Ranking individuals by `S_between` ranks
their groups, which is a different statement.

⚠️ **The rows are the participants of the *grouped* set, not of the fit.**
Anyone the grouping does not cover is not in the calculation at all, so the
numbers are conditional on that set:

```{r acc_ind_rows, eval=have_acc}
length(unique(bi$id))          # participants in the decomposition
nrow(acc$Y)                    # participants in the fit
sum(is.na(stage))              # not covered by the grouping
```

That is a change of question, not a defect. A grouping covering fewer people
gives different contributions because it is a different conditioning set.

`summary()` prints the leading participants for every query, next to the
per-block table they decompose:

```{r acc_ind_summary, eval=have_acc}
summary(sens_acc)
```

And the whole distribution is easier seen than tabulated:

```{r acc_ind_plot, eval=have_acc, fig.wide=TRUE, fig.height=9, dpi=120, fig.cap="One panel per query. Bars are stacked by assay, and the pale bar at the foot of each panel is everyone outside the top ten."}
plot(sens_acc, type = "individuals", top_n = 10)
```

**How to read this figure.** One bar per participant, stacked by assay, so you
see who carries the sensitivity and which assay it reaches them through at the
same time. Compare the real query against its noise controls: a query whose
sensitivity rests on a handful of people looks very different from one spread
evenly across the cohort, and that difference is invisible in `S_total` alone.

**The pale "all others" bar is the point of the figure, not decoration.** A
top-N chart without the remainder silently overstates concentration -- ten tall
bars look like the whole story whether they carry most of the total or a tenth
of it. Drawn together, the comparison is immediate:

```{r acc_ind_conc, eval=have_acc}
d   <- bi[bi$query == "stage_score", ]
per <- sort(rowsum(d$S_total, d$id)[, 1], decreasing = TRUE)
# `top_n` only changes how many bars are named; the remainder bar absorbs the
# rest either way, so the panel always accounts for the full total.
round(c(top10 = sum(head(per, 10)), all_others = sum(per[-(1:10)])) / sum(per), 3)
```

⚠️ **Participants absent from a block contribute a numerical zero there, not an
exact one.** `miniACC` is full of them -- most participants are missing from at least
one assay -- and their contribution to that assay comes back at the scale of
rounding error rather than at `0`:

```{r acc_ind_absent, eval=have_acc}
range(d$S_total[d$S_total < 1e-20])
```

They are deliberately not rounded down: the value is defined, it simply happens
to be negligible, and thresholding it would hide the difference between "absent"
and "present and unimportant".

**What these per-participant numbers are, and what they are not.** They are an
**exploratory** decomposition of the first-order sensitivity diagnostic. The
block-level quantity they decompose has empirical calibration evidence, with a
MIXTO verdict; **no individual-level calibration has been established**, and the
block-level evidence does not transfer to the individual level on its own. They
are sensitivity contributions -- **not** variance, not uncertainty, not causal
influence, and not a guarantee of correctness or reliability for any participant.
A large bar says the fitted alignment is more sensitive to that participant's
measurements under this configuration; it says nothing about that person, and it
does not rank people in any way that survives a change of fit, ridge, query,
block or grouping.

Read this view as a **diagnostic pointer**: if one or two participants carry a
large share of a finding, that is worth looking at -- an outlying sample, a
processing anomaly -- before the finding is reported as a property of the cohort.

# Would I get the same answer with a different sample? {#is-the-subspace-reproducible}

`mgcca_stability()` refits on resamples and measures how far the subspace moves.
`"subsample"` needs no grouping; `"loco"` leaves out one level of `group` at a
time -- the generic form of leave-one-cohort-out.

Two columns come back, and the second is the one that gets misread.

`overlap` is the agreement with the reference subspace, computed **on the
participants both fits contain**. One means the same directions; lower means they
moved.

`rank_margin` is *not* a measure of how well separated the components are. It is
the smallest singular value of the reference frame once restricted to the shared
participants, and it asks whether the comparison is well posed at all: if the
frame becomes near-degenerate under restriction, a high overlap is comparing
directions barely present in the shared set and means very little. An orthonormal
frame has every singular value equal to one, so a value well below one is the
warning, not the reassurance.

Both outcomes are worth seeing, so this section shows one of each rather than
picking the flattering one.

## A result you can lean on

The cardiovascular fit has 646 individuals. Holding out 10% of them at a time:

```{r cardio_stab}
stab_ok <- mgcca_stability(fit, B = 30, fraction = 0.90, seed = 1)
stab_ok
```

```{r cardio_stab_read, echo=FALSE, results='asis'}
s1 <- stab_ok$summary_metrics
cat(sprintf(paste0(
  "Worst case %.4f over %d resamples, with a rank margin of %.4f. Read that as: ",
  "had you recruited a somewhat different set of individuals, you would have ",
  "recovered the same axes, and the comparison itself was well posed throughout. ",
  "This is what it looks like when a result does not depend on who happened to be ",
  "sampled.\n"),
  s1$overlap_min, s1$n_resamples, s1$rank_margin_min))
```

## A result you cannot

```{r acc_stab, eval=have_acc}
stab <- mgcca_stability(acc, method = "both", group = stage,
                        B = 40, fraction = 0.632, seed = 1)
stab
```

```{r acc_stab_read, eval=have_acc, echo=FALSE, results='asis'}
sm <- stab$summary_metrics
ss <- sm[sm$method == "subsample", ]; lo <- sm[sm$method == "loco", ]
# The verdict is DERIVED from the number, never printed beside it. A canned
# sentence that says "essentially the same axes" whatever the minimum turns out
# to be is not a reading: it is decoration sitting next to data, and sooner or
# later it contradicts the figure above it.
verdict <- function(x) {
  if (x >= 0.99) return("the axes are effectively unchanged")
  if (x >= 0.95) return("the axes move only slightly")
  if (x >= 0.80) return("the axes move noticeably, so individual directions should be interpreted with care")
  "the axes are NOT stable here, and individual directions should not be interpreted"
}
loco_line <- if (lo$overlap_min >= 0.95) {
  "no single group is holding the result up on its own."
} else {
  "at least one group matters a good deal, and dropping it moves the axes appreciably."
}
cat(sprintf(paste0(
  "Here the median is %.4f but the worst case is %.4f. At the median, %s. ",
  "**The worst case is the number that matters**, and at %.4f, %s.\n\n",
  "Leaving out an entire stage group at a time gives a worst case of %.4f: %s\n\n",
  "This is not a defect of the data. `miniACC` has 80 patients, and a third of ",
  "them removed at once is a large perturbation for a sample that size. The point ",
  "is that the diagnostic *said so* instead of returning a comfortable number.\n"),
  ss$overlap_median, ss$overlap_min, verdict(ss$overlap_median),
  ss$overlap_min, verdict(ss$overlap_min), lo$overlap_min, loco_line))
```

```{r acc_stab_plot, eval=have_acc, fig.height=4.2}
plot(stab)
```

Plotting overlap against rank margin makes the caution explicit: the points to
distrust are the ones on the left, whatever their height.

## What made the difference

The difference is not that one data set is good and the other bad. It is how hard
each was pushed, and how much sample there was to absorb it.

```{r sweep, echo=FALSE}
sw <- do.call(rbind, lapply(c(0.632, 0.80, 0.90, 0.95), function(f) {
  m <- mgcca_stability(fit, B = 20, fraction = f, seed = 1)$summary_metrics
  data.frame(fraction = f, worst_overlap = round(m$overlap_min, 4),
             median_overlap = round(m$overlap_median, 4),
             worst_rank_margin = round(m$rank_margin_min, 4))
}))
knitr::kable(sw, caption = "Cardiovascular fit: the same data at four resampling severities.")
```

Two things to take from this table.

The first is that **stability is a property of the data *and* of the question you
ask of it**. Reporting an overlap without the fraction it was computed at says
almost nothing.

The second is the useful one: **the rank margin falls alongside the overlap**.
When the resamples get small enough for the axes to move, the reference frame
also becomes ill-conditioned on the shared participants -- so the diagnostic
warns you that the comparison is degrading at the same time as it degrades. You
are not left to guess whether a low overlap means real instability or an
ill-posed comparison; the two columns together tell you.

# Practical matters

## Where the computation happens

`backend = "auto"` decides by block size, exactly as `mgcca(route = "auto")`
does, but `"memory"` and `"hdf5"` force the choice. The backend that ran is
recorded in the result, so a number is never ambiguous about how it was produced.

```{r backend, eval=FALSE}
mgcca_sensitivity(acc, Q, group = stage, backend = "hdf5")
```

`block_size` forces the out-of-core block size. Its normal value is `0` (let
`BigDataStatMeth` choose); a small positive value exists so the blocked path can
be exercised on small data, which is how the package tests it.

## Keeping your results

Persistence is opt-in. An analysis function should not write to your file unasked,
particularly when the fit may be shared or read-only.

```{r store, eval=FALSE}
sens_acc <- mgcca_sensitivity(acc, Q, group = stage,
                              store = TRUE, store_name = "stage_sensitivity")
```

Results go under `RELIABILITY/<name>/` in the fit's own HDF5 file, with the same
attribute manifest the package uses elsewhere. An existing name is **refused**
rather than overwritten: two runs with different settings under one name would
leave a manifest that no longer describes the data beside it, and nothing
downstream could detect that.

## What you can and cannot claim

These are first-order sensitivity coefficients of a fitted operator, under the
scaling, weighting and availability conventions of the fit that produced them.
They support statements about *where a query's alignment is sensitive*, and about
*whether a subspace is reproducible*. They do not support causal statements, and
a high `R_between` does not by itself mean an association is an artefact -- it
means the alignment and the grouping move together in this fit.

When the retained and discarded parts of the spectrum are not separated, the
first-order expansion is not meaningful at all. The result carries a validity flag
and the print method says so rather than letting the numbers stand unqualified.

# Session information

```{r session}
sessionInfo()
```
